\bf Longitudinal Principal Manifold Estimation

### 2024-07-10 paper/main.tex
Updates from Overleaf

Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curve in the structure at the upper-left corner of each plot. Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME. It also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with high levels of curvature.

@@

The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data. To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same set of individuals described previously. Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data. Considering the same individual shown in Figure \ref{fig:lhipp_cross_sections}, cross sectional results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}. The cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

@@

A summary of the between-visit variability in the volume estimates of the subcortical structures considered here is shown in Table \ref{table:adni_volume_sds}. In addition to reporting the standard deviations of the volume estimates, we also report regression-adjusted estimates of variability to account for cases in which the structure cannot be assumed to be constant over time, as is the case for those with AD. The regression-adjusted summaries estimate the standard deviation of the residuals of a simple linear regression of volume on time. Meanwhile, in cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e. raw data, LPME, and PME). We then took the mean of these values for each estimate source. Table \ref{table:adni_volume_sds} shows that PME yields the smoothest volume estimates for both hippocampuses. This conclusion aligns with the visual evidence of inconsistency in the LPME estimates caused by the irregular shape of the hippocampus. However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Structure & Regression-Adjusted & Data & LPME & PME  \\
    \hline
    Left Hippocampus & No & 129 & 132 & \textbf{106} \\
    Right Hippocampus & No & 137 & 136 & \textbf{110} \\
    Left Hippocampus & Yes & 100 & 108 & \textbf{87.1} \\
    Right Hippocampus & Yes & 108 & 121 & \textbf{85.9} \\
    Left Thalamus & No & 57.5 & \textbf{31.3} & 42.0 \\
    Right Thalamus & No & 108 & \textbf{66.4} & 85.3 \\
    Left Thalamus & Yes & 45.8 & \textbf{20.6} & 35.4 \\
    Right Thalamus & Yes & 89.3 & \textbf{50.1} & 73.1 \\
    \hline
  \end{tabular}
  \caption{{\footnotesize Mean of the empirical standard deviation of volume estimates computed for each ADNI participant. Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.} }
  \label{table:adni_volume_sds}
\end{table}

>>>>

Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curves in the structure near the top and bottom corners of each plot. Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME. It also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with high levels of curvature.

@@

The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data. To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same set of individuals described previously. Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data. Considering the same individual shown in Figure \ref{fig:lhipp_cross_sections}, cross sectional results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figure \ref{fig:lthal_cross_sections}. The cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus, though the LPME estimate still falls inside the observed surface.

@@

A summary of the between-visit variability in the volume estimates of the subcortical structures considered here is shown in Table \ref{table:adni_volume_sds}. In addition to reporting the standard deviations of the volume estimates, we also report regression-adjusted estimates of variability to account for cases in which the structure cannot be assumed to be constant over time, as is the case for those with AD. The regression-adjusted summaries estimate the standard deviation of the residuals of a simple linear regression of volume on time. Meanwhile, in cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e. raw data, LPME, and PME). We then took the median of these values for each estimate source. Table \ref{table:adni_volume_sds} shows that, despite the visual evidence of inconsistency in the LPME estimates of the hippocampus caused by its irregular shape, LPME still yields the smoothest volume estimates for both hippocampuses. LPME also results in the smoothest volume estimates when applied to the thalamus, showing even greater improvements over the alternative volume estimates.

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Structure & Regression-Adjusted & Data & LPME & PME  \\
    \hline
    Left Hippocampus & No & 129 & \textbf{119} & 124 \\
    Right Hippocampus & No & 133 & \textbf{124} & 179 \\
    Left Hippocampus & Yes & 104 & \textbf{93.2} & 98.4 \\
    Right Hippocampus & Yes & 107 & \textbf{100} & 149 \\
    Left Thalamus & No & 55.7 & \textbf{28.5} & 42.5 \\
    Right Thalamus & No & 108 & \textbf{61.7} & 93.3 \\
    Left Thalamus & Yes & 45.2 & \textbf{16.0} & 35.3 \\
    Right Thalamus & Yes & 86.9 & \textbf{42.7} & 76.6 \\
    \hline
  \end{tabular}
  \caption{{\footnotesize Median of the empirical standard deviation of volume estimates computed for each ADNI participant. Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time.} }
  \label{table:adni_volume_sds}
\end{table}

### 2024-07-10 paper/supplement.tex
Updates from Overleaf

@@

\begin{figure}
\centering
\subfloat{
\label{fig:adni_lhipp_result}
\subfloat[\centering Data]{
\includegraphics[height=10cm]{adni_plots/adni_lhipp_data_plot}
}
\hfill
\subfloat[\centering LPME]{
\includegraphics[height=10cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
}
\hfill
\subfloat[\centering PME]{
\includegraphics[height=10cm]{adni_plots/adni_lhipp_pme_plot}
}
}
\hfill
\subfloat{
\label{fig:adni_lthal_result}
\subfloat[\centering Data]{
\includegraphics[height=10cm]{adni_plots/adni_lthal_data_plot}
}
\hfill
\subfloat[\centering LPME]{
\includegraphics[height=10cm]{adni_plots/adni_lthal_lpme_isomap_plot}
}
\hfill
\subfloat[\centering PME]{
\includegraphics[height=10cm]{adni_plots/adni_lthal_pme_plot}
}
}
\caption{Left Hippocampus (left), and Left Thalamus (right), Cognitive Healthy ADNI Participant. The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue. The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
\label{fig:adni_result}
\end{figure}

>>>>

\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}

\setcounter{figure}{0}
\setcounter{table}{0}

@@

\begin{figure}

  \begin{subfigure}{\textwidth}
    \label{fig:adni_lhipp_result}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lhipp_data_plot}
      \caption{Data}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lhipp_lpme_isomap_plot}
      \caption{LPME}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lhipp_pme_plot}
      \caption{PME}
    \end{subfigure}   
  \end{subfigure}

  \begin{subfigure}{\textwidth}
    \label{fig:adni_lthal_result}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lthal_data_plot}
      \caption{Data}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lthal_lpme_isomap_plot}
      \caption{LPME}
    \end{subfigure}
    \begin{subfigure}{\textwidth}
      \centering
      \includegraphics[height=3cm]{adni_plots/adni_lthal_pme_plot}
      \caption{PME}
    \end{subfigure}   
  \end{subfigure}

  \caption{Left Hippocampus (panels a-c), and Left Thalamus (panels d-f), Cognitive Healthy ADNI Participant. The raw surface data, displayed in red, show slight changes in orientation that are absent in the LPME estimates, shown in blue. The estimates obtained by the PME and LPME algorithms appear unable to accurately capture the true shape of the hippocampus at points with high levels of curvature.}
  \label{fig:adni_result}
\end{figure}
